---
title: "HW02: Basics of Data Frames"
subtitle: "Stat 133, Fall 2017"
author: "Woo Sik (Lewis) Kim"
output: github_document
---
**Importing the data set in different ways; first, using "read.csv()", then "read_csv()":**

• Using "read.csv()", and displaying the resulting data frame:

```{r}
dataset1 <- read.csv(file = "data/nba2017-player-statistics.csv",
                     header = TRUE,
                     colClasses = c("character",
                     "character",
                     "factor", "character", "double",
                     rep("integer", 19)))
str(dataset1)
```

• Using "read_csv()", and displaying the resulting data frame (I will be using the data set from read_csv() for the remainder of the homework):

```{r}
require(readr)
library(dplyr)
dataset2 <- read_csv('data/nba2017-player-statistics.csv',
                     col_types =
                       list(
                         Player = col_character(),
                         Team = col_character(),
                         Position = col_factor(c('C', 'PF','PG','SF','SG')),
                         Experience = col_character(),
                         Salary = col_double(),
                         Rank = col_integer(),
                         Age = col_integer(),
                         GP = col_integer(),
                         GS = col_integer(),
                         MIN = col_integer(),
                         FGM = col_integer(),
                         FGA = col_integer(),
                         Points3 = col_integer(),
                         Points3_atts = col_integer(),
                         Points2 = col_integer(),
                         Points2_atts = col_integer(),
                         FTM = col_integer(),
                         FTA = col_integer(),
                         OREB = col_integer(),
                         DREB = col_integer(),
                         AST = col_integer(),
                         STL = col_integer(),
                         BLK = col_integer(),
                         TO = col_integer()
                         ))
str(dataset2)
```

**Changing all instances of "R" (character) to "0" (integer) in the column "Experience":**

```{r}
dataset2[dataset2$Experience == "R", "Experience"] <- 0
dataset2$Experience <- as.integer(dataset2$Experience)
```

**Computing and adding new columns to dataset2 to calculate EFF:**

```{r}
Missed_FG <- dataset2$FGA - dataset2$FGM
Missed_FT <- dataset2$FTA - dataset2$FTM
PTS <- 2*dataset2$Points2 + 3*dataset2$Points3 + dataset2$FTM
REB <- dataset2$OREB + dataset2$DREB
MPG <- dataset2$MIN / dataset2$GP

dataset2 <- cbind(dataset2, Missed_FG, Missed_FT, PTS, REB, MPG)

EFF <- (dataset2$PTS + dataset2$REB + dataset2$AST + dataset2$STL + 
          dataset2$BLK - dataset2$Missed_FG - dataset2$Missed_FT - 
          dataset2$TO) / dataset2$GP

dataset2 <- cbind(dataset2, EFF)
```

**Computing summary statistics for efficiency values, and creating a histogram:**

```{r}
# Declaring dataset2$EFF as a new variable "Efficiency" for histogram title(s).
Efficiency <- dataset2$EFF 

summary(Efficiency)
hist(Efficiency)
```

**Processing the EFF data to get the top 10 players, and their associated name, team, salary, and EFF values:**

```{r}
desc_eff <- select(dataset2, Player, Team, Salary, EFF)
head(arrange(desc_eff, desc(dataset2$EFF)), 10)
```

**Obtaining player(s) with negative EFF value(s):**

```{r}
neg_eff <- filter(select(dataset2, Player, EFF), dataset2$EFF < 0)
str(neg_eff)
```

• Top 10 players in descending EFF order: "Russell Westbrook," "James Harden," "Anthony Davis," "LeBron James," "Karl-Anthony Towns," "Kevin Durant," "Giannis Antetokounmpo," "DeMarcur Cousins," "Jimmmy Butler," "Hassan Whiteside."

• Player(s) with a negative EFF value: "Patricio Garino: -0.6"

**Calculating the correlation coefficients between EFF and all the variables used in the EFF formula (in order of PTS, REB, STL, AST, BLK, Missed_FT, Missed_FG, TO), and adding them to the vector "corr_coeff":**

```{r}
corr_coeff <- c(cor(dataset2$EFF, dataset2$PTS), 
                cor(dataset2$EFF,dataset2$REB),
                cor(dataset2$EFF, dataset2$STL), 
                cor(dataset2$EFF, dataset2$AST),
                cor(dataset2$EFF, dataset2$BLK),
                -1*cor(dataset2$EFF, dataset2$Missed_FT),
                -1*cor(dataset2$EFF, dataset2$Missed_FG),
                -1*cor(dataset2$EFF, dataset2$TO))

sorted_eff <- sort(corr_coeff, decreasing = TRUE)
```

**Creating a barchart with the correlations:**

```{r fig.width = 10, fig.height = 5}
barplot(sorted_eff, main = "Correlations between Player
        Stats and EFF", beside=TRUE,
        ylim = c(-1, 1), names = c("PTS", "REB", "STL", "AST", "BLK",
                                   "Missed_FT", "Missed_FG", "TO"),
        col = c("grey", "grey", "grey","grey", "grey", "red", "red", "red"))
abline(0, 0)
```

**Creating a scatterplot of "Efficiency" vs. "Salary" using the entire NBA data set (dataset2):**

```{r}
plot(dataset2$EFF, dataset2$Salary, xlab = "Efficiency", ylab = "Salary")
lines(lowess(dataset2$EFF, dataset2$Salary), col = "blue")
```

Also, compute the linear correlation coefficient between them. What can you
say about the relationship between these two variables?

**Computing the linear correlation coefficient between "Efficiency" and "Salary" for ALL players:**

```{r}
cor(dataset2$EFF, dataset2$Salary)
```

• The relationship between efficiency and salary for all players does not seem to be very strong. There is a positive linear relationship between the two variables (as one increases, the other also increases), but there may not be statistical significance to their correlation. This also has to do with the fact that we're not looking at enough variables, since this correlation only considers efficiency statistic vs. salary.

**Selecting players who have MPG values of more than 20, and creating data frame
"players2":**

```{r}
players2 <- select(filter(dataset2, dataset2$MPG > 20), Player, Salary, EFF)
str(players2)
```
                          
**Creating a scatterplot between "Efficiency" and "Salary" using the data frame "players2" (players whose MPG values > 20):**

```{r}
plot(players2$EFF, players2$Salary, xlab = "Efficiency", ylab = "Salary")
lines(lowess(players2$EFF, players2$Salary), col = "blue")
```

**Computing the linear correlation coefficient between "Efficiency" and "Salary" for players whose MPG values > 20:**

```{r}
cor(players2$EFF, players2$Salary)
```

• The correlation coefficient between efficiency and salary for the "more established players" seem to be lower than the coefficient for all players by about 22% (0.655 / 0.536), meaning there is a weaker correlation for the "more established players" than among everyone. This may be because there are more rookies, and they contribute more to the EFF statistic simply due to sheer numbers. It also may be that there are other variables we're not considering.

**Comments and Reflections (Answered in Order)**

• The hardest part of this class so far has been getting used to R's syntax, and the different style of programming it entails compared to things like Python and Java.

• Combining functions such as filter() and select(), and using "new" functions such as sort() from online documentations was easy.

• Yes, I needed help from office hours with debugging and properly creating charts/graphs.

• This homework took about 3 hours to complete (spread out over multiple days).

• The most time consuming part was properly importing the .csv data set.

• No, I think I understood everything fully once I learned how to do it.

• Yes, some syntactical errors caused frustration.

• No, not necessarily for this homework. Although, organizing "unorganized" NBA data and visualizing them was pretty great.
