---
title: "HW03: Ranking Teams"
subtitle: "Stat 133, Fall 2017"
author: "Woo Sik (Lewis) Kim"
output: github_document
---
**Source of the Data:**
  
```{r}
source('/Users/Lewis/Stat\ 133/stat133-hws-fall17/hw03/code/make-teams-data.R')
```

**Generating a Salary Plot Using GGPlot2:**
  
```{r}
salary_plot <- ggplot(data = teams, aes(x =reorder(team, salary) ,y =salary, group = 1)) + labs(x= "Teams", y = "Salary") +
  geom_bar(stat='identity') + coord_flip()

salary_plot <- salary_plot + geom_hline(yintercept = mean(teams$salary),color = "red")

salary_plot
```

**Generating a Points Plot Using GGPlot2:**
  
```{r}
points_plot <- ggplot(data = teams, aes(x =reorder(team, points) ,y =points, group = 1)) + labs(x= "Teams", y = "Points") +
  geom_bar(stat='identity') + coord_flip() 

points_plot <- points_plot + geom_hline(yintercept = mean(teams$points),color = "red")

points_plot
```

**Generating an Efficiency Plot Using GGPlot2:**
  
```{r}
eff_plot <- ggplot(data = teams, aes(x =reorder(team, efficiency) ,y =efficiency, group = 1)) + labs(x= "Teams", y = "Points") +
  geom_bar(stat='identity') + coord_flip()

eff_plot <- eff_plot + geom_hline(yintercept = mean(teams$efficiency),color = "red")

eff_plot
```

**Performing a "Principal Components Analysis" (PCA) on the following variables (refer to data dictionary for definitions): points3, points2, free_throws, off_rebounds, assists, steals, blocks, turnovers, fouls.**
  
```{r}
pca_teams <- select(teams, points3, points2, free_throws, off_rebounds, def_rebounds, assists, steals, blocks, turnovers, fouls)

pca_teams <- prcomp(pca_teams, scale.=TRUE)
pca_teams
```

**Creating a data frame with the eigenvales from PCA:**
  
```{r}
eigenvalues = pca_teams$sdev^2
prop = round(pca_teams$sdev^2 / sum(pca_teams$sdev^2), 4)
cprop = cumsum(prop)
evals <- data.frame(eigenvalues, prop, cprop)

evals
```

**Using the first 2 PCs to generate a scatter plot of the teams:**

```{r}
pc1 <- pca_teams$x[,1] 
pc2 <- pca_teams$x[,2] 

plot(pc1, pc2, main = "PCA Scatter Plot (PC1 and PC2)", cex = 0)

text(pc1, pc2, labels=teams$team)
```

**Building an Index Based on PC1 by transforming PC1 through Rescaling (Involves Calculating z1 and s1 scores):**

```{r}
s1 <- 100 * ((pc1 - min(pc1)) / max(pc1) - min(pc1))

pc1_plot <- ggplot(data = teams, aes(x =reorder(teams$team, s1) ,y =s1, group = 1)) + labs(x= "Team", y = "PC1 (scaled from 0 to 100)") +
  geom_bar(stat='identity') + coord_flip() 

pc1_plot
```
